Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
s/cleartext/plaintext/
Christina Boura:
  https://simple.wikipedia.org/wiki/Plaintext
  https://simple.wikipedia.org/wiki/Cleartext

%% * Executive summary of modern symmetric cryptography
%% ** iterate pure `round` function N times
%% *** thread a `state`
%% **** initially: clear text
%% **** finally: cipher text
%% *** mixing in constants derived from a secret key
%% **** computed off-line
%% **** off the critical path

For our present purposes, the essence of modern symmetric
cryptographic primitives can be distilled in a few paragraph. A
primitive can be understood as a purely functional program. It takes
two inputs: the ``cleartext'' and a list of ``round subkeys'' (which
are derived from a single key through a non-performance-critical
process called ``key schedule'' that we shall ignore here). The output
is the ``ciphertext''. The cleartext, the ciphertext and each
individual round subkeys are processed as ``blocks'' of binary data.
The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).

Looking at the process of turning the cleartext into a ciphertext, one
realizes that cryptographers have been programming in the State monad
all along: ciphers are described as operating on a ``state'',
initialized to be the cleartext and read off at the end as the
ciphertext. This state is updated by a ``round'' function that is
repeated a given number of times (of the order of ten iterations). The
exact number of iterations is a trade-off between latency (less
iterations means faster computation) and security (more iterations
makes for more costly cryptanalysis effort). The round function itself
is built compositionally from 3 components: a key mixing layer (making
the encryption process reversible), a confusion layer (smoothing local
correlations among the inputs) and a diffusion layer (spreading
correlations across the whole output block).

>>>>

%% * Executive summary of modern symmetric cryptography
%% ** iterate pure `round` function N times
%% *** thread a `state`
%% **** initially: plain text
%% **** finally: cipher text
%% *** mixing in constants derived from a secret key
%% **** computed off-line
%% **** off the critical path

For our present purposes, the essence of modern symmetric
cryptographic primitives can be distilled in a few paragraph. A
primitive can be understood as a purely functional program. It takes
two inputs: the ``plaintext'' and a list of ``round subkeys'' (which
are derived from a single key through a non-performance-critical
process called ``key schedule'' that we shall ignore here). The output
is the ``ciphertext''. The plaintext, the ciphertext and each
individual round subkeys are processed as ``blocks'' of binary data.
The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).

Looking at the process of turning the plaintext into a ciphertext, one
realizes that cryptographers have been programming in the State monad
all along: ciphers are described as operating on a ``state'',
initialized to be the plaintext and read off at the end as the
ciphertext. This state is updated by a ``round'' function that is
repeated a given number of times (of the order of ten iterations). The
exact number of iterations is a trade-off between latency (less
iterations means faster computation) and security (more iterations
makes for more costly cryptanalysis effort). The round function itself
is built compositionally from 3 components: a key mixing layer (making
the encryption process reversible), a confusion layer (smoothing local
correlations among the inputs) and a diffusion layer (spreading
correlations across the whole output block).
